Phenotypes_Planting_2019

### 2020-01-06 data-raw/phenotypes_2019.Rmd
Plots of 2019 phenotyping errors and 2019 phenotypes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(XLConnect)
library(rlang)
library(switchgrassGWAS)
library(cowplot)
library(viridis)
library(bigsnpr)
library(scales)

@@

@@

```
Anything where DEAD_2018 = 2 is called as "Not virgatum" in the past, so should be dropped. Does Jason agree?

@@

```{r}
load("../data/Taxa.rda")

phenotypes_wide_2019 <- phemean_2019 %>%
  gather(FRZ_DMG:SPADCOR_158, key = "Phenotype", value = value) %>%
  filter(!is.na(value)) %>%
  # saveRDS(file = "Cleaned pvdiv Phenotypes from 2018 for 784g 2019-05-16.rds")
  unite(col = "SITE_PHE", SITE, Phenotype) %>%
  spread(key = SITE_PHE, value = value) %>%
  right_join(Taxa)
```

@@

sites$Factor <- factor(sites$Factor, levels = c("KING", "PKLE", "TMPL", "OVTN", "STIL", "CLMB", "LINC", "FRMI", "KBSM", "BRKG"))
sites$Factor <- factor(sites$Factor, levels = c("BRKG", "KBSM", "FRMI", "LINC", "CLMB", "STIL", "OVTN", "TMPL", "PKLE", "KING"))
levels(sites$Factor)

phenotypes_long_2019$SITE <- factor(phenotypes_long_2019$SITE, levels = c("KING", "PKLE", "TMPL", "OVTN", "STIL", "CLMB", "LINC", "FRMI", "KBSM", "BRKG"))
phenotypes_long_2019$SITE <- factor(phenotypes_long_2019$SITE, levels = c("BRKG", "KBSM", "FRMI", "LINC", "CLMB", "STIL", "OVTN", "TMPL", "PKLE", "KING"))
phenotypes_long_2019$PHE <- factor(phenotypes_long_2019$PHE, levels = c("GR1", "GR50", "GR100", "EM1", "EM50", "FL1", "FL50", "TC_EOS", "HT_PAN_EOS", "FRZ_DMG", "CHLR_101", "CHLR_121", "CHLR_136", "CHLR_JASON_138", "CHLR_150.", "SPAD_158", "SPADCOR_158"))

@@

phenotypes_long_2019 %>%
  filter(!(grepl("CHLR", PHE)) & !(grepl("SPAD", PHE))) %>%
  ggplot(aes(x = Value)) + 
  geom_histogram(aes(fill = SITE)) + 
  facet_wrap(~PHE, scales = "free")

library(lubridate)
phenology_long_2019 %>%
  mutate(Days = as_date(Value)) %>%
  ggplot(aes(x = Days)) + 
  geom_histogram(aes(fill = SITE)) + 
  facet_grid(SITE~PHE, scales = "free") +
  scale_fill_viridis(discrete = TRUE, end = 0.95, direction = -1) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Date") + ylab("Number of Individuals") +
  geom_vline(xintercept = c(120), linetype = 2) +
  geom_vline(xintercept = c(181), linetype = 3, color = "grey20") +
  geom_vline(xintercept = c(243), linetype = 4, color = "grey80")

@@

load("../data/metadata.rda")

metadata <- metadata %>%
  mutate(Genetic_subpop = case_when(
           is.na(JL_genetic_subpop) ~ "Unknown",
           JL_genetic_subpop == "other" ~ "Admixed",
           TRUE ~ JL_genetic_subpop
           ))
metadata$Genetic_subpop <- factor(metadata$Genetic_subpop, levels = c("TX", "GC,TX", "GC", "SEC", "NEC,SEC", "NEC", "MW", "Admixed", "Unknown"))
metadata$Ecotype <- factor(metadata$Ecotype, levels = c("TX", "Coast", "Up"))

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(XLConnect)
library(rlang)
library(switchgrassGWAS)
library(cowplot)
library(viridis)
library(bigsnpr)
library(scales)
library(lubridate)

@@

phenos_2019 %>%
  filter(SRV == "W" & DEAD_2018 == 2)

@@

```
Anything where DEAD_2018 = 2 is called as "Not virgatum" in the past, so should be dropped. Does Jason agree?
FRMI	F2610	2610	J203	J203.A	NA	156	NA	NA	W	Likely not virgatum, Weak	switchgrass?
FRMI	F2707	2707	J421	J421.A	NA	149	176	NA 	W	Likely not virgatum, Weak	switchgrass?

@@

```{r}
phenotypes_wide_2019 <- phemean_2019 %>%
  gather(FRZ_DMG:SPADCOR_158, key = "Phenotype", value = value) %>%
  filter(!is.na(value)) %>%
  # saveRDS(file = "Cleaned pvdiv Phenotypes from 2018 for 784g 2019-05-16.rds")
  unite(col = "SITE_PHE", SITE, Phenotype) %>%
  spread(key = SITE_PHE, value = value) %>%
  right_join(Taxa)
```

@@

#sites$Factor <- factor(sites$Factor, levels = c("KING", "PKLE", "TMPL", "OVTN", "STIL", "CLMB", "LINC", "FRMI", "KBSM", "BRKG"))
#sites$Factor <- factor(sites$Factor, levels = c("BRKG", "KBSM", "FRMI", "LINC", "CLMB", "STIL", "OVTN", "TMPL", "PKLE", "KING"))
#levels(sites$Factor)

# phenotypes_long_2019$SITE <- factor(phenotypes_long_2019$SITE, levels = c("KING", "PKLE", "TMPL", "OVTN", "STIL", "CLMB", "LINC", "FRMI", "KBSM", "BRKG"))
phenotypes_long_2019$SITE <- factor(phenotypes_long_2019$SITE, levels = c("BRKG", "KBSM", "FRMI", "LINC", "CLMB", "STIL", "OVTN", "TMPL", "PKLE", "KING"))
phenotypes_long_2019$PHE <- factor(phenotypes_long_2019$PHE, levels = c("GR1", "GR50", "GR100", "EM1", "EM50", "FL1", "FL50", "TC_EOS", "HT_PAN_EOS", "FRZ_DMG", "CHLR_101", "CHLR_121", "CHLR_136", "CHLR_JASON_138", "CHLR_150.", "SPAD_158", "SPADCOR_158"))

@@

```{r}
phenotypes_long_2019 %>%
  filter(!(grepl("CHLR", PHE)) & !(grepl("SPAD", PHE))) %>%
  ggplot(aes(x = Value)) + 
  geom_histogram(aes(fill = SITE)) + 
  facet_wrap(~PHE, scales = "free")

phenology_long_2019 %>%
  mutate(Days = as_date(Value)) %>%
  ggplot(aes(x = Days)) + 
  geom_histogram(aes(fill = SITE)) + 
  facet_grid(SITE~PHE, scales = "free") +
  scale_fill_viridis(discrete = TRUE, end = 0.95, direction = -1) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Date") + ylab("Number of Individuals") +
  geom_vline(xintercept = c(120), linetype = 2) +
  geom_vline(xintercept = c(181), linetype = 3, color = "grey20") +
  geom_vline(xintercept = c(243), linetype = 4, color = "grey80")

@@

```{r}
metadata <- metadata %>%
  mutate(Genetic_subpop = case_when(
           is.na(JL_genetic_subpop) ~ "Unknown",
           JL_genetic_subpop == "other" ~ "Admixed",
           TRUE ~ JL_genetic_subpop
           ),
        Ecotype = case_when(
          Ecotype == "TX" ~ "Texas",
          Ecotype == "Coast" ~ "Coastal",
          Ecotype == "Up" ~ "Upland"
          ))
metadata$Genetic_subpop <- factor(metadata$Genetic_subpop, levels = c("TX", "GC,TX", "GC", "SEC", "NEC,SEC", "NEC", "MW", "Admixed", "Unknown"))
metadata$Ecotype <- factor(metadata$Ecotype, levels = c("Texas", "Coastal", "Upland"))

@@

## Other phenotypes by Ecotype/Subpop

```{r}
phenotypes_long_2019 %>%
  filter(grepl("TC_EOS", PHE) | grepl("HT_PAN_EOS", PHE) | grepl("FRZ", PHE)) %>%
  left_join(metadata, by = "PLANT_ID") %>%
  filter(!(Genetic_subpop %in% c("Admixed", "Unknown")) & 
           PLANT_ID != "AP13") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(aes(fill = Genetic_subpop)) + 
  facet_grid(SITE~PHE, scales = "free") +
  scale_fill_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Value") + ylab("Count of Individuals")

save_plot("2019_Phenotypic_Data_by_Site_and_Genetic_Subpop_2019-12-30.png", plot = last_plot(), base_height = 10, base_width = 9)

phenotypes_long_2019 %>%
  filter(grepl("TC_EOS", PHE) | grepl("HT_PAN_EOS", PHE) | grepl("FRZ", PHE)) %>%
  left_join(metadata, by = "PLANT_ID") %>%
  filter(!(Genetic_subpop %in% c("Unknown")) & 
           PLANT_ID != "AP13") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(aes(fill = Ecotype)) + 
  facet_grid(SITE~PHE, scales = "free") +
  scale_fill_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("Value") + ylab("Count of Individuals")

save_plot("2019_Phenotypic_Data_by_Site_and_Ecotype_2019-12-30.png", plot = last_plot(), base_height = 10, base_width = 9)
```

# Do dates make sense for each SITE*PLANT_ID?

In the aggregate these values seem to make sense, check them for individual cases.

```{r}
phenotypes_long_2019 %>%
  filter(grepl("TC_EOS", PHE) | grepl("HT_PAN_EOS", PHE)) %>%
  left_join(metadata, by = "PLANT_ID") %>%
  filter(!(Genetic_subpop %in% c("Unknown")) & 
           PLANT_ID != "AP13") %>%
  pivot_wider(names_from = PHE, values_from = Value) %>%
  ggplot(aes(x = TC_EOS, y = HT_PAN_EOS)) +
  geom_point(aes(color = Ecotype), alpha = 0.2) +
  facet_wrap(~SITE) +
  scale_color_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right")

phe_comp <- phenology_long_2019 %>%
  left_join(metadata, by = "PLANT_ID") %>%
  filter(!(Genetic_subpop %in% c("Unknown")) & 
           PLANT_ID != "AP13") %>%
  pivot_wider(names_from = PHE, values_from = Value)
```

```{r}
phe_comp %>%
  ggplot(aes(x = GR50, y = EM50)) +
  geom_point(aes(color = Ecotype), alpha = 0.2) +
  facet_wrap(~SITE) +
  scale_color_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right") + 
  geom_abline(slope = 1, intercept = 0)
save_plot("Greenup_vs_Emergence_50percent_by_Site_2019-12-31.png", plot = last_plot(), base_height = 4)
phe_comp %>%
  ggplot(aes(x = GR50, y = FL50)) +
  geom_point(aes(color = Ecotype), alpha = 0.2) +
  facet_wrap(~SITE) +
  scale_color_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right") + 
  geom_abline(slope = 1, intercept = 0)
save_plot("Greenup_vs_Flowering_50percent_by_Site_2019-12-31.png", plot = last_plot(), base_height = 4)
phe_comp %>%
  ggplot(aes(x = EM50, y = FL50)) +
  geom_point(aes(color = Ecotype), alpha = 0.2) +
  facet_wrap(~SITE) +
  scale_color_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right") + 
  geom_abline(slope = 1, intercept = 0)
save_plot("Emergence_vs_Flowering_50percent_by_Site_2019-12-31.png", plot = last_plot(), base_height = 4)
```

TMPL has some EM1 and EM50 issues. not collected very much at later timepoints.
OVTN: I don't have EM or FL data. Was this data not collected?

TMPL and PKLE (and KING, and STIL less so) have a huge spread of FL1 vs GR1. Also true for FL1 vs GR100 and FL50 vs GR100. GxE?!

phe_comp %>%
  ggplot(aes(x = FL1, y = FL50)) +
  geom_point(aes(color = Ecotype), alpha = 0.5) +
  facet_wrap(~SITE) +
  scale_color_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right") + 
  geom_abline(slope = 1, intercept = 0)
save_plot("Possible_Flowering_Errors_below_the_line_by_Site_2019-12-30.png", plot = last_plot(), base_height = 4)

phe_comp %>%
  ggplot(aes(x = GR1, y = GR100)) +
  geom_point(aes(color = Ecotype), alpha = 0.2) +
  #facet_wrap(~SITE) +
  scale_color_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right") + 
  geom_abline(slope = 1, intercept = 0)

save_plot("Possible_Greenup_Errors_below_the_line_by_Site_2019-12-30.png", plot = last_plot(), base_height = 4)

phe_comp %>%
  ggplot(aes(x = EM1, y = EM50)) +
  geom_point(aes(color = Ecotype), alpha = 0.4) +
  facet_wrap(~SITE) +
  scale_color_viridis(discrete = TRUE, end = 0.8, direction = -1) +
  theme(legend.position = "right") + 
  geom_abline(slope = 1, intercept = 0)
save_plot("Possible_Emergence_Errors_below_the_line_by_Site_2019-12-30.png", plot = last_plot(), base_height = 4)
```
Generally, EM50 should not be less than EM1, and it certainly shouldn't be much less (like where EM1 is 200 and EM50 is 50, must be an error)
```{r}
phe_comp %>%
  mutate(EMdiff = EM50 - EM1) %>%
  filter(EMdiff < 0) %>%
  dplyr::select(SITE:PLANT_ID, EM1, EM50, EMdiff)

phe_comp %>%
  mutate(FLdiff = FL50 - FL1) %>%
  filter(FLdiff < 0) %>%
  dplyr::select(SITE:PLANT_ID, FL1, FL50, FLdiff)
```
SITE PLOT_GL PLOT_LC ACC PLANT_ID EM1 EM50 EMdiff

STIL	S1110	1110	J587	J587.A	186	185	-1
STIL	S2907	2907	J073	J073.A	190	184	-6
STIL	S3603	3603	J210	J210.A	188	186	-2
TMPL	T1417	1417	J321	J321.A	182	180	-2
TMPL	T2101	2101	J586	J586.B	199	182	-17
TMPL	T2103	2103	J008	J008.A	182	126	-56
TMPL	T2111	2111	J326	J326.B	199	19	-180
TMPL	T2203	2203	J614	J614.B	142	141	-1
TMPL	T2411	2411	J270	J270.A	199	99	-100

For GxE study:

Greenup date (50%)
time to bolting (EM50 - GR50)
Panicle emergence date (50%)
time to flowering (FL50 - GR50)
Flowering date (50%)
Tiller count EOS 2019
Panicle height EOS 2019

Also GR1, GR100, EM1, FL1, CHLR/SPAD in KING, FRZ_DAM, but the 7 above seem more interesting.

# Remove outliers

```{r}
phe_2019_fil <- phenotypes_long_2019 %>%
  left_join(metadata, by = "PLANT_ID") %>%
  filter(!is.na(JL_genetic_subpop)) %>%
  pivot_wider(names_from = PHE, values_from = Value) %>%
  dplyr::select(SITE:PLANT_ID, LIB_PHENO, Ecotype, GR1:FRZ_DMG) %>%
  filter(LIB_PHENO == "Y") %>%
  mutate(EM50 = ifelse(EM50 - EM1 < 0, NA, EM50),
         FL50 = ifelse(FL50 - FL1 < 0, NA, FL50),
         TBOLT = EM50 - GR50,
         TFLOW = FL50 - GR50)

phemean_2019 <- phe_2019_fil %>%
  group_by(PLANT_ID, SITE) %>%
  summarise(FRZ_DMG = mean(FRZ_DMG, na.rm = TRUE),
            GR1 = mean(GR1, na.rm = TRUE),
            GR50 = mean(GR50, na.rm = TRUE),
            GR100 = mean(GR100, na.rm = TRUE),
            EM1 = mean(EM1, na.rm = TRUE),
            EM50 = mean(EM50, na.rm = TRUE),
            FL1 = mean(FL1, na.rm = TRUE),
            FL50 = mean(FL50, na.rm = TRUE),
            TC_EOS = mean(TC_EOS, na.rm = TRUE),
            HT_PAN_EOS = mean(HT_PAN_EOS, na.rm = TRUE),
            CHLR_101 = mean(CHLR_101, na.rm = TRUE),
            CHLR_121 = mean(CHLR_121, na.rm = TRUE),
            CHLR_136 = mean(CHLR_136, na.rm = TRUE),
            CHLR_JASON_138 = mean(CHLR_JASON_138, na.rm = TRUE),
            CHLR_150. = mean(CHLR_150., na.rm = TRUE),
            SPAD_158 = mean(SPAD_158, na.rm = TRUE),
            SPADCOR_158 = mean(SPADCOR_158, na.rm = TRUE),
            TBOLT = mean(TBOLT, na.rm = TRUE),
            TFLOW = mean(TFLOW, na.rm = TRUE)
  ) %>%
  mutate_all( ~ case_when(!is.nan(.x) ~ .x)) %>% # if the column is not NA, keep the value; else replace with NA as nothing is provided as another choice in case_when.
  filter(!is.na(SITE) & !is.na(PLANT_ID))
```

```{r}
phenotypes_wide_2019 <- phemean_2019 %>%
  dplyr::select(PLANT_ID:SITE, GR1:HT_PAN_EOS, TBOLT, TFLOW) %>%
  gather(c(GR1:TFLOW), key = "Phenotype", value = value) %>%
  filter(!is.na(value)) %>%
  # saveRDS(file = "Cleaned pvdiv Phenotypes from 2018 for 784g 2019-05-16.rds")
  unite(col = "SITE_PHE", SITE, Phenotype) %>%
  spread(key = SITE_PHE, value = value) %>%
  right_join(Taxa)

save(phenotypes_wide_2019, file = "../data/phenotypes_2019.rda")
```

# How many plants are phenotyped at each site?

```{r}
colSums(!is.na(phenotypes_wide_2019))

phenotypes_wide_2019 %>%
  left_join(metadata) %>%
  group_by(Ecotype, is.na(BRKG_GR50)) %>%
  tally()

phenotypes_long_2019 %>%
  left_join(metadata) %>%
  group_by(SITE, PHE, Ecotype) %>%
  tally()
phemean_2019 %>%
  left_join(metadata) %>%
  dplyr::select(PLANT_ID, SITE, GR1:HT_PAN_EOS) %>%
  ggplot(aes(x = PLANT_ID, y = SITE)) +
  geom_heatm
  pivot_longer(cols = GR1:HT_PAN_EOS, names_to = "PHE", values_to = "Value") %>%
  group_by(SITE, PHE) %>%
  tally()

tibble(PLANT_ID = Taxa$PLANT_ID, PHE = rowSums(!is.na(phenotypes_wide_2019)) - 1) %>%
  filter(PHE > 0)

colremove <- names(which(colSums(!is.na(phenotypes_wide_2019)) < 200))

phe_gwas_2019 <- phenotypes_wide_2019 %>%
  dplyr::select(-colremove, -ends_with("1"), -ends_with("100"))

plotphe <- phe_gwas_2019 %>%
  #dplyr::select(PLANT_ID, ends_with("TBOLT")) %>%
  pivot_longer(cols = BRKG_EM50:TMPL_TFLOW, names_to = "PHE", values_to = "VALUE") %>%
  left_join(metadata) %>%
  filter(!is.na(VALUE)) %>%
  separate(PHE, into = c("SITE", "_", "PHE"), sep = c(4,5))

plotphe$SITE <- factor(plotphe$SITE, levels = c("BRKG", "KBSM", "FRMI", "LINC", "CLMB", "STIL", "OVTN", "TMPL", "PKLE", "KING"))
plotphe$PHE <- factor(plotphe$PHE, levels = c("GR50", "EM50", "FL50", "TBOLT", "TFLOW", "TC_EOS", "HT_PAN_EOS"))

plotphe %>%
  group_by(PHE, Ecotype) %>% 
  ggplot(aes(x = SITE)) +
  geom_bar(aes( fill = Ecotype)) +
  facet_wrap(~PHE) +
  scale_fill_viridis(discrete = TRUE, end = 0.95, direction = -1) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, 
                                                              hjust = 1)) +
  xlab("Site") + ylab("Number of Individuals")

save_plot("Number_of_Individuals_per_Phenotype_2_2020-01-04.png", plot = last_plot(), base_height = 4*2.5, base_width = 5*2.5)

summary(metadata$Ecotype)
```
